First Order Alternation

### 2018-07-09 quantifiers/draft.tex
ok

In the upcoming developments it is more convenient to work with
logical formulae defining executions. Let $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the rest of
this session. For any $i \in \nat$, we denote by $\stamp{Q}{i} =
\set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} = \set{\stamp{x}{i}
  \mid x \in X}$ the sets of time-stamped predicates and variables,
respectively. For a formula $\psi$ and $i \in \nat$, we define
$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
formula in which all input variables and state predicates are replaced
by their time-stamped counterparts. Given a sequence of input events
$\alpha = a_1 \ldots a_n$, the \emph{path formula} of $\alpha$ in $\A$
is defined as follows:
\[\pathform{\A}{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow \stamp{\psi}{i}\]
To formalize the relation between the low-level configuration-based
execution semantics and the symbolic path formulae, consider a word
$w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any execution forest $\F$ of $\A$
over $w$ is associated an interpretation $\I_{\F}$ of the set of
time-stamped predicates $\set{\stamp{q}{i} \mid q \in Q, i \in [n]}$,
defined as follows:
\[\I_{\F}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}}
  \mid q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in
    $\F$}}\enspace.\]

\begin{lemma}\label{lemma:path-formula}
  For any word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have: 
  \[\minsem{\pathform{\A}{w_\Sigma}}_{w_\Data} = \set{\I_{\F} \mid \F
    \text{ is an execution of $\A$ over $w$}}\enspace.\]
\end{lemma}
\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
  $\I,w_\Data \models \pathform{\A}{w_\Sigma}$. We show that there
  exists an execution $\F$ of $\A$ over $w$ such that $\I = \I_\F$, by
  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
  $\pathform{\A}{w_\Sigma} = \stamp{\iota}{0}$. Because $\iota$ is a
  sentence, the valuation $w_\Data$ is not important in $\I,w_\Data
  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
  $\J,u_\Data \models \pathform{\A}{u_\Sigma}$ and $\J$ is moreover
  minimal. By the induction hypothesis, there exists an execution $\G$
  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$ be
  the subformula of $\pathform{\A}{w_\Sigma}$ corresponding to the
  application(s) of the transition rule $q(y_1,\ldots,y_{\#(q)})
  \arrow{a_n}{} \psi$ at the $(n-1)$-th step. Let $\nu$ be the
  valuation that assigns each $y_i$ the value $d_i$, for all $i \in
  [1,\#(q)]$ and behaves like $w_\Data$ outside of
  $\set{y_1,\ldots,y_{\#(d)}}$. Because $I, w_\Data \models \forall
  y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$,
  we have $\I \in \sem{\stamp{\psi}{n}}_{w_\Data}$ and let $\K$ be one
  of the minimal interpretations such that $\K \subseteq \I$ and $\K
  \in \sem{\stamp{\psi}{n}}_{w_\Data}$. It is not hard to see that
  $\K$ exists and is unique, otherwise we could take the pointwise
  intersection of two or more such interpretations. We define the
  interpretation $\bar{\K}(q) = \bar{\K}(\stamp{q}{n})$ for all $q \in
  Q$. We have that $\bar{\K} \in \minsem{\psi}_{\nu}$ --- if
  $\bar{\K}$ was not minimal, $\K$ was minimal to start with,
  contradiction. Then we extend the execution $\G$ by appending to
  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
  the cube $\cube{\bar{\K}}$. Clearly the result is an execution of
  $\A$ over $w$.

  ``$\supseteq$'' \comment[ri]{todo} \qed}

>>>>

In the upcoming developments it is more convenient to work with
logical formulae defining executions. Let $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the rest of
this session. For any $i \in \nat$, we denote by $\stamp{Q}{i} =
\set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} = \set{\stamp{x}{i}
  \mid x \in X}$ the sets of time-stamped predicates and variables,
respectively. For a formula $\psi$ and $i \in \nat$, we define
$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
formula in which all input variables and state predicates (and only
those symbols) are replaced by their time-stamped counterparts. Given
a sequence of input events $\alpha = a_1 \ldots a_n$, the \emph{path
  formula} of $\alpha$ (in $\A$) is defined as follows:
\[\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow \stamp{\psi}{i}\]
The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
be clear from the context. To formalize the relation between the
low-level configuration-based execution semantics and the symbolic
path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
execution forest $\F$ of $\A$ over $w$ is associated an interpretation
$\I_{\F}$ of the set of time-stamped predicates $\set{\stamp{q}{i}
  \mid q \in Q, i \in [n]}$, defined as follows:
\[\I_{\F}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}}
  \mid q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in
    $\F$}}\enspace.\]

\begin{lemma}\label{lemma:path-formula}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have:
  \[\minsem{\pathform{w_\Sigma}}_{w_\Data} = \set{\I_{\F} \mid \F
    \text{ is an execution of $\A$ over $w$}}\enspace.\]
\end{lemma}
\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
  $\I,w_\Data \models \pathform{w_\Sigma}$. We show that there exists
  an execution $\F$ of $\A$ over $w$ such that $\I = \I_\F$, by
  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
  $\pathform{w_\Sigma} = \stamp{\iota}{0}$. Because $\iota$ is a
  sentence, the valuation $w_\Data$ is not important in $\I,w_\Data
  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
  $\J,u_\Data \models \pathform{u_\Sigma}$ and $\J$ is moreover
  minimal. By the induction hypothesis, there exists an execution $\G$
  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$ be
  the subformula of $\pathform{w_\Sigma}$ corresponding to the
  application(s) of the transition rule $q(y_1,\ldots,y_{\#(q)})
  \arrow{a_n}{} \psi$ at the $(n-1)$-th step. Let $\nu$ be the
  valuation that assigns each $y_i$ the value $d_i$, for all $i \in
  [1,\#(q)]$ and behaves like $w_\Data$ outside of
  $\set{y_1,\ldots,y_{\#(d)}}$. Because $\I, w_\Data \models \forall
  y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$,
  we have $\I \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of
  the minimal interpretations such that $\K \subseteq \I$ and $\K \in
  \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
  and is unique, otherwise we could take the pointwise intersection of
  two or more such interpretations. We define the interpretation
  $\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in
  Q$. We have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
  $\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
  contradiction. Then we extend the execution $\G$ by appending to
  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
  the cube $\cube{\overline{\K}}$. By repeating this step for all
  leaves of a tree in $\G$, we obtain an execution of $\A$ over $w$.

  ``$\supseteq$'' Let $\F$ be an execution of $\A$ over $w$. We show
  that $\I_{\F}$ is a minimal interpretation such that $\I_{\F},
  w_\Data \models \pathform{w_\Sigma}$, by induction on $n \geq
  0$. For $n=0$, $\F$ is a cube from $\cube{\minsem{\iota}}$, by
  definition. Then $\I_{\F} \models \stamp{\iota}{0}$ and moreover, it
  is a minimal such interpretation. For the inductive case $n > 0$,
  let $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\G$ be the
  restriction of $\F$ to $u$. Consequently, $\I_\G$ is the restriction
  of $\I_\F$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$. By the inductive
  hypothesis, $\I_{\G}$ is a minimal interpretation such that
  \(\I_{\G}, u_{\Data} \models \pathform{u_\Sigma}\). Since
  $\I_{\F}(\stamp{q}{n}) = \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
    q(d_1,\ldots,d_{\#(q)}) \text{ occurs on the $n$-th level in
      $\F$}}$, we have $\I_{\F}, w_\Data \models \varphi$, for each
  subformula $\varphi = \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$ of
  $\pathform{w_\Sigma}$, by the execution semantics of $\A$. This is
  the case because the children of each node labeled with
  $q(d_1,\ldots,d_{\#(q)})$ on the $(n-1)$-th level of $\F$ form a
  cube from $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation
  that assigns each $y_i$ the value $d_i$ and behaves like $w_\Data$,
  otherwise. Now supppose, for a contradiction, that $\I_{\F}$ is not
  minimal and let $\J \subsetneq \I_{\F}$ be an interpretation such
  that $\J,w_\Data \models \pathform{w_\Sigma}$. First, we show that
  the restriction $\J'$ of $\J$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$
  must coincide with $\I_{\G}$. Assuming this is not the case,
  i.e.\ $\J' \subsetneq \I_{\G}$, contradicts the minimality of
  $\I_{\G}$. Then the only possibility is that $\J(\stamp{q}{n})
  \subsetneq \I_{\F}(\stamp{q}{n})$, for some $q \in Q$. Let
  $p_1(y_1,\ldots,y_{\#(p_1)}) \arrow{a_n}{} \psi_1, \ldots,
  p_k(y_1,\ldots,y_{\#(p_k)}) \arrow{a_n}{} \psi_k$ be the set of
  transition rules in which the predicate symbol $q$ occurs. Then it
  must be the case that, for some node $p_i(d_1,\ldots,d_{\#(p_i)})$
  on the $(n-1)$-th level of $\G$, the set of children does not form a
  minimal cube from $\cube{\minsem{\stamp{\psi_i}{n}}}$, which
  contradicts the execution semantics of $\A$. \qed}
